---
title: "Penalized Factor Analysis: grid search"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Penalized Factor Analysis: grid search}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 600px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
options(width = 100)
```

## Introduction

**Aim**. This vignette shows how to fit a penalized factor analysis model with
the scad and mcp penalties using the routines in the `penfa` package. The
employed penalty is aimed at encouraging sparsity in the factor loading matrix.
Since the scad and mcp penalties cannot be used with the automatic tuning
procedure (see Geminiani et al., 2021 for details), the optimal tuning parameter
will be found through grid searches.

**Data**. For illustration purposes, we use the cross-cultural data set `ccdata`
containing the standardized ratings to 12 items concerning organizational
citizenship behavior. Employees from different countries were asked to rate
their attitudes towards helping other employees and giving suggestions for
improved work conditions. The items are thought to measure two latent factors:
**help**, defined by the first seven items (`h1` to `h7`), and **voice**,
represented by the last five items (`v1` to `v5`). See `?ccdata` for details.

This data set is a standardized version of the one in the
[`ccpsyc`](https://github.com/Jo-Karl/ccpsyc/) package, and only considers
employees from Lebanon and Taiwan (i.e., `"LEB"`, `"TAIW"`). 
This vignette is meant as a demo of the capabilities of `penfa`; please refer to 
Fischer et al. (2019) and Fischer and Karl (2019) for a description and
analysis of these data.

Let us load and inspect `ccdata`.

```{r dataset}
library(penfa)
data(ccdata)

summary(ccdata)
```

## Model specification

Before fitting the model, we need to write a model syntax describing the
relationships between the items and the latent factors. To facilitate its
formulation, the rules for the syntax specification broadly follow the ones
required by [lavaan](https://CRAN.R-project.org/package=lavaan/). The
syntax must be enclosed in single quotes `' '`.

```{r syntax}
syntax = 'help  =~   h1 + h2 + h3 + h4 + h5 + h6 + h7 + 0*v1 + v2 + v3 + v4 + v5
          voice =~ 0*h1 + h2 + h3 + h4 + h5 + h6 + h7 +   v1 + v2 + v3 + v4 + v5'
```


The factors `help` and `voice` appear on the left-hand side, whereas the
observed variables on the left-hand side. Following the rationale in Geminiani
et al. (2021), we only specify the minimum number of identification constraints.
We are setting the scales of the factors by fixing their factor variances to 1.
This can be done in one of two ways: 1) by adding `'help ~~ 1*help'` and `'voice
~~ 1*voice'`  to the syntax above; or 2) by setting the argument `std.lv = TRUE`
in the fitting function (see below). To avoid rotational freedom, we fix one
loading per factor to zero. Parameters can be easily fixed to user-defined
values through the pre-multiplication mechanism. By default, unique variances
are automatically added to the model, and the factors are allowed to correlate.
These specifications can be modified by altering the syntax (see `?penfa` for
details on how to write the model syntax).

## Model fitting

The core of the package is given by the `penfa` function, a short form for
*PENalized Factor Analysis*, that implements the framework discussed in
Geminiani et al. (2021). If users decide for either the
`scad` or `mcp` penalties, they need to use `strategy = "fixed"` since the
automatic procedure is not feasible.

### Scad 

We start off with the scad. In the function call, we now specify `pen.shrink =
"scad"`, and we provide through the `eta` argument the fixed value of the tuning
parameter to be employed during optimization (here, for instance, 0.05). The
name given to the starting value (here, the factor loading matrix `"lambda"`)
reflects the parameter matrix to be penalized. All of its elements are
penalized, which means here that the penalization is applied to all factor
loadings (except the ones fixed for identification). The scad penalty relies on
an additional shape parameter, which is set by default to 3.7  (Fan and Li
2001). This value can be conveniently modified through the `a.scad` argument.
See `?penfaOptions` for additional details on the possible options.

```{r scad}
scad.fit <- penfa(## factor model
                  model  = syntax,
                  data   = ccdata,
                  std.lv = TRUE,
                  ## penalization
                  pen.shrink = "scad",
                  eta = list(shrink = c("lambda" = 0.05), diff = c("none" = 0)),
                  ## fixed tuning
                  strategy = "fixed")
```

#### Grid search

In order to find the optimal value of the tuning parameter, a grid search needs
to be conducted, and the optimal model is the one with the lowest GBIC
(Generalized Bayesian Information Criterion). For demo purposes, we use a grid
of 51 values evenly spaced between 0 and 0.15. However, for accurate analyses,
please consider finer grids (of at least 200 elements) with an upper bound
reasonable for the data at hand.

```{r scad_grid_search}
# Grid of values for tuning parameter
eta.grid <- seq(0, 0.15, length.out = 51)

# Return GBIC from a converged and admissible penfa model with fixed tuning
penfa.fixedTun <- function(eta, penalty, ...){
  
  fitted <- penfa(model = syntax, data = ccdata, 
                  std.lv = TRUE, pen.shrink = penalty, 
                  eta = list(shrink = c("lambda" = eta), diff = c("none" = 0)),
                  strategy = "fixed", verbose = FALSE, ...)
  
  if(all(fitted@Vcov$solution) & fitted@Vcov$admissibility)
    return(BIC(fitted))
}

# additional penfaOptions can be passed
GBIC.scad <- sapply(eta.grid, penfa.fixedTun, penalty = "scad") 
```

The optimal tuning parameter is the one generating the penalized model with the
lowest `GBIC`.

```{r scad_plot, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
optimtun.scad <- eta.grid[[which.min(GBIC.scad)]]

# To plot GBIC across tuning values
# p <- plotly::plot_ly(x = eta.grid, y = GBIC.scad, type = 'scatter', mode = 'lines')
# plotly::layout(p, xaxis = list(showline = TRUE),
#               title = list(text = "GBIC values across tuning parameters"))
```


We can ultimately fit the model with the optimal tuning parameter (here, `r optimtun.scad`). 
The `summary` method details information on the model characteristics, the
optimization and penalization procedures as well as the parameter estimates with
associated standard errors and confidence intervals. The *Type* column
distinguishes between the *fixed* parameters set to specific values for
identification, the *free* parameters that have been estimated through ordinary
maximum likelihood, and the penalized (*pen*) parameters. The standard errors
here have been computed as the square root of the inverse of the penalized
Fisher information matrix (Geminiani et al., 2021). The last columns report 95%
confidence intervals (CI) for the model parameters. Standard errors and CI of
the penalized parameters shrunken to zero are not displayed. A different
significance level can be specified through the `level` argument in the
`summary` call.


```{r scad_fit}
scad.fit <- penfa(## factor model 
                  model = syntax, 
                  data = ccdata, 
                  std.lv = TRUE, 
                  ## penalization
                  pen.shrink = "scad", 
                  # optimal tuning
                  eta = list(shrink = c("lambda" = optimtun.scad), diff = c("none" = 0)),
                  strategy = "fixed", 
                  verbose = FALSE)
summary(scad.fit)
```

The penalty matrix can be inspected and plotted as shown in
["plotting-penalty-matrix"](https://egeminiani.github.io/penfa/articles/articles/plotting-penalty-matrix.html).

#### Implied moments

The implied moments (here, the covariance matrix) can be found via the `fitted`
method.

```{r scad_implied}
implied <- fitted(scad.fit)
implied
```

### Factor scores

Lastly, the factor scores can be calculated via the `penfaPredict` function.

```{r fscores}
fscores <- penfaPredict(scad.fit)
head(fscores)
```

### Mcp 

We can fit a penalized factor model with the mcp penalty in a way similar to the
scad. By default the shape parameter of the mcp is set to 3. This value can be
conveniently modified through the `a.mcp` argument. See `?penfaOptions` for
additional details on the possible options.

```{r mcp_grid_search, fig.height=4, fig.width=6, message=FALSE, warning=FALSE}
GBIC.mcp <- sapply(eta.grid, penfa.fixedTun, penalty = "mcp")

optimtun.mcp <- eta.grid[[which.min(GBIC.mcp)]]
optimtun.mcp
```

The tuning value equal to `r optimtun.mcp` generated the model with the lowest GBIC.

```{r}
mcp.fit <- penfa(## factor model 
                 model = syntax, 
                 data = ccdata, 
                 std.lv = TRUE, 
                 ## penalization
                 pen.shrink = "mcp", 
                 # optimal tuning
                 eta = list(shrink = c("lambda" = optimtun.mcp), diff = c("none" = 0)),
                 strategy = "fixed", 
                 verbose = FALSE)
summary(mcp.fit)
```


## Conclusion

Implementing the above approach in the multiple-group case would imply carrying
out grid searches in three dimensions to find the optimal tuning parameter
vector. This is clearly not advisable, as it would introduce further
complications and possibly new computational problems and instabilities. For
this reason, we suggest users to rely on the automatic tuning parameter
selection procedure, whose applicability is demonstrated in the
package vignettes for single (`vignette("automatic-tuning-selection")`) and
multiple-group (`vignette("multiple-group-analysis")`) penalized
models.


## R Session

```{r}
sessionInfo()
```



## References

* Fan, J., & Li, R. 2001. "Variable Selection via Nonconcave Penalized Likelihood 
and Its Oracle Properties." Journal of the American Statistical 
Association 96(456), 1348–60. [https://doi.org/10.1198/016214501753382273](https://doi.org/10.1198/016214501753382273)

* Fischer, R., Ferreira, M. C., Van Meurs, N. et al. (2019). "Does
Organizational Formalization Facilitate Voice and Helping Organizational
Citizenship Behaviors? It Depends on (National) Uncertainty Norms." Journal of
International Business Studies, 50(1), 125-134.
[https://doi.org/10.1057/s41267-017-0132-6](https://doi.org/10.1057/s41267-017-0132-6)

* Fischer, R., & Karl, J. A. (2019). "A Primer to (Cross-Cultural) Multi-Group 
Invariance Testing Possibilities in R." Frontiers in psychology, 10, 1507. [https://doi.org/10.3389/fpsyg.2019.01507](https://doi.org/10.3389/fpsyg.2019.01507)

* Geminiani, E. (2020). "A Penalized Likelihood-Based Framework for Single and
Multiple-Group Factor Analysis Models." PhD thesis, University of Bologna.
[http://amsdottorato.unibo.it/9355/](http://amsdottorato.unibo.it/9355/)

* Geminiani, E., Marra, G., & Moustaki, I. (2021). "Single- and Multiple-Group
Penalized Factor Analysis: A Trust-Region Algorithm Approach with Integrated
Automatic Multiple Tuning Parameter Selection." Psychometrika, 86(1), 65-95. [https://doi.org/10.1007/s11336-021-09751-8](https://doi.org/10.1007/s11336-021-09751-8)
