---
title: "Plotting Penalty Matrices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Plotting Penalty Matrices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
pre {
  max-height: 600px;
  overflow-y: auto;
}

pre[class] {
  max-height: 600px;
}
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Introduction

**Aim**. This vignette shows how to plot the penalty matrices from the single-
and multiple-group penalized factor models estimated in 
`vignette("automatic-tuning-selection")` and `vignette("multiple-group-analysis")`.

**Data**. For illustration purposes, we use the cross-cultural data set `ccdata`
containing the standardized ratings to 12 items concerning organizational
citizenship behavior. Employees from different countries were asked to rate
their attitudes towards helping other employees and giving suggestions for
improved work conditions. The items are thought to measure two latent factors:
**help**, defined by the first seven items (`h1` to `h7`), and **voice**,
represented by the last five items (`v1` to `v5`). See `?ccdata` for details.

This data set is a standardized version of the one in the
[`ccpsyc`](https://github.com/Jo-Karl/ccpsyc/) package, and only considers
employees from Lebanon and Taiwan (i.e., `"LEB"`, `"TAIW"`). 
This vignette is meant as a demo of the capabilities of `penfa`; please refer to 
Fischer et al. (2019) and Fischer and Karl (2019) for a description and
analysis of these data.


Let us load and inspect `ccdata`.

```{r dataset, R.options = list(width = 100)}
library(penfa)
data(ccdata)

summary(ccdata)
```


## Penalized factor analysis

Let us fit the penalized factor model with alasso and automatic tuning procedure 
as described in `vignette("automatic-tuning-selection")`.

```{r model_fit}
# Specify syntax
syntax = 'help  =~   h1 + h2 + h3 + h4 + h5 + h6 + h7 + 0*v1 + v2 + v3 + v4 + v5
          voice =~ 0*h1 + h2 + h3 + h4 + h5 + h6 + h7 +   v1 + v2 + v3 + v4 + v5'

# Get adaptive weights
mle.fit <- penfa(## factor model 
                 model = syntax, 
                 data  = ccdata,
                 std.lv = TRUE,
                 ## (no) penalization
                 pen.shrink = "none",
                 eta = list(shrink = c("none" = 0), diff = c("none" = 0)),
                 strategy = "fixed",
                 verbose = FALSE)
mle.weights <- coef(mle.fit)


# Model fit
alasso.fit <- penfa(## factor model
                    model  = syntax,
                    data   = ccdata,
                    std.lv = TRUE,
                    ## penalization
                    pen.shrink = "alasso",
                    eta = list(shrink = c("lambda" = 0.01), diff = c("none" = 0)),
                    ## automatic procedure
                    strategy = "auto",
                    gamma = 4,
                    ## alasso
                    weights = mle.weights,
                    verbose = FALSE)
alasso.fit
```

The penalty matrix can be extracted through the `penmat` function.

```{r}
alasso_penmat <- penmat(alasso.fit)
```


Applying the `plot` method to the penalty matrix allows us to visualize an
interactive heatmap of the log of the absolute value of the estimated penalty
matrix. Due to space constraints, it may occur that some of the parameter labels
on the axes are hidden. If it is the case, users can zoom in on the area of
interest and inspect the corresponding penalty values.

```{r plotly_heatmap, fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
plot(alasso_penmat)
```



## Penalized Multiple-Group Factor Analysis

Let us fit the penalized multiple-group factor model with alasso and automatic 
multiple tuning procedure as described in `vignette("multiple-group-analysis")`.


```{r}
# Specify syntax
syntax.mg = 'help  =~ 1*h1 +          h2 +          h3 + h4 + h5 + h6 + h7 + 0*v1 + v2 + v3 + v4 + v5
             voice =~ 0*h1 + start(0)*h2 + start(0)*h3 + h4 + h5 + h6 + h7 + 1*v1 + v2 + v3 + v4 + v5
             
             h2 + h3 + h4 + h5 + h6 + h7 + v2 + v3 + v4 + v5 ~ 1 
             help  ~ NA*1
             voice ~ NA*1 '


# Get adaptive weights
mle.fitMG <- penfa(## factor model
                   model = syntax.mg,
                   data  = ccdata,
                   group = "country",
                   ## (no) penalization
                   pen.shrink = "none",
                   pen.diff = "none",
                   eta = list(shrink = c("lambda" = 0), diff = c("none" = 0)),
                   strategy = "fixed",
                   verbose = FALSE) 
mle.weightsMG <- coef(mle.fitMG)

# Model fit
alasso.fitMG <- penfa(## factor model
                      model = syntax.mg,
                      data = ccdata,
                      group = "country",
                      int.lv.free = TRUE,
                      ## penalization
                      pen.shrink = "alasso",
                      pen.diff = "alasso",
                      eta = list(shrink = c("lambda" = 0.01),
                                 diff   = c("lambda" = 0.1, "tau" = 0.01)),
                      ## automatic procedure
                      strategy = "auto",
                      gamma = 4,
                      ## alasso
                      weights = mle.weightsMG)

```


The complete penalty matrix is stored in `alasso.fit@Penalize@Sh.info$S.h`, but
it can be easily extracted via the `penmat` function. This matrix is the sum of
the penalty matrices for Penalty 1 (`sparsity.penmat`), Penalty 2
(`loadinvariance.penmat`), and Penalty 3 (`intinvariance.penmat`). Unique
variances, factor (co)variances and factor means were not affected by the
penalization, so their entries in the penalty matrices are equal to zero.

Through the `plot` method, we can visualize an interactive heatmap of the log of
the absolute value of each estimated penalty matrix (because of the wide element
range). Due to space constraints given by the large number of parameters, some
parameter labels on the axes may be hidden. In this case, users can zoom in on
the area of interest and inspect the penalty values for the parameters of
interest.

```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
full.penmat <- penmat(alasso.fitMG)
plot(full.penmat)
```

The above penalty matrix is the sum of the following three individuals penalty
matrices.

#### Sparsity

This penalty matrix shrinks the small factor loadings of each group to zero.
Apart from the group loadings, all the remaining entries of the penalty matrix
are equal to zero.

```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
sparsity.penmat <- penmat(alasso.fitMG, type = "shrink", which = "lambda")
plot(sparsity.penmat)
```

#### Loading invariance

This penalty matrix shrinks the pairwise group differences of the factor
loadings to zero.

```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
loadinvariance.penmat <- penmat(alasso.fitMG, type = "diff", which = "lambda")
plot(loadinvariance.penmat)
```

#### Intercept invariance

This penalty matrix shrinks the pairwise group differences of the intercepts to
zero.

```{r fig.height=6, fig.width=8, message=FALSE, warning=FALSE}
intinvariance.penmat <- penmat(alasso.fitMG, type = "diff", which = "tau")
plot(intinvariance.penmat)
```


## R Session

```{r}
sessionInfo()
```



## References

* Fischer, R., Ferreira, M. C., Van Meurs, N. et al. (2019). "Does
Organizational Formalization Facilitate Voice and Helping Organizational
Citizenship Behaviors? It Depends on (National) Uncertainty Norms." Journal of
International Business Studies, 50(1), 125-134.
[https://doi.org/10.1057/s41267-017-0132-6](https://doi.org/10.1057/s41267-017-0132-6)

* Fischer, R., & Karl, J. A. (2019). "A Primer to (Cross-Cultural) Multi-Group 
Invariance Testing Possibilities in R." Frontiers in psychology, 10, 1507. [https://doi.org/10.3389/fpsyg.2019.01507](https://doi.org/10.3389/fpsyg.2019.01507)

* Geminiani, E. (2020). "A Penalized Likelihood-Based Framework for Single and
Multiple-Group Factor Analysis Models." PhD thesis, University of Bologna.
[http://amsdottorato.unibo.it/9355/](http://amsdottorato.unibo.it/9355/)

* Geminiani, E., Marra, G., & Moustaki, I. (2021). "Single- and Multiple-Group
Penalized Factor Analysis: A Trust-Region Algorithm Approach with Integrated
Automatic Multiple Tuning Parameter Selection." Psychometrika, 86(1), 65-95. [https://doi.org/10.1007/s11336-021-09751-8](https://doi.org/10.1007/s11336-021-09751-8)
